---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(picohdr)
```

# picohdr

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![CRAN](https://www.r-pkg.org/badges/version/picohdr)](https://CRAN.R-project.org/package=picohdr)
[![R-CMD-check](https://github.com/coolbutuseless/picohdr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/picohdr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{picohdr}` is a package for reading and writing HDR (High Dynamic Range) images. 
Supported image types are types [`PFM`](https://www.pauldebevec.com/Research/HDR/PFM/) 
and [`EXR`](https://openexr.com/).  Tools for tone-mapping HDR images and adjusting array values are also included.

* `PFM` *Portable Float Map* is a very simple 2D and 3D array serialization format
  that is used to store floating point data for images.
* `EXR` is a more comprehensive HDR image standard, but `picohdr` only supports a commonly occurring feature subset:
    * Single-part scanline images only
    * Compression modes: `NONE`, `ZIP`, `ZIPS` (`ZIPS` support is for read only)
    * Parsing meta-information works on all image types (single-part, scanline, 
      tiled, deep, and multi-part images)

The package prefix is `pico` as it only addresses a small subset of the EXR image specification.

## What's in the box

* PFM support
    * `read_pfm()`, `write_pfm()` for handling *Portable Float Map* (PFM) images
* EXR support
    * `read_exr()` for reading single-part scanline EXR images. Supports `NONE`, `ZIP` or `ZIPS`
      compression modes. 
    * `write_exr()` for writing single-part scanline EXR images. Supports `NONE` and `ZIP`
      compression modes. 
    * `exr_info()` return information about an EXR file.  Supports single-part, 
      multi-part, tiled and deep images.
* Tone mapping for converting from HDR to LDR (low dynamic range) raster image.
  These functions work hard to shift array values into the range [0,1] with 
  the aim to approximate the look of the HDR image in a more limited medium. 
    * `tm_reinhard()`  
    * `tm_reinhard_basic()`           
    * `tm_reinhard_variant()` 
* General array adjustments
    * `adj_clamp()` clamp values outside the given limits
    * `adj_gamma()` gamma correction
    * `adj_infinite()` remove infinite values by setting them to the largest/smallest finite values
    * `adj_rescale()` linearly rescale all values into the given range
    * `adj_shift_negatives_global()` find the most negative value in the data and 
       shift all values up by this amount so that there are no negatives.
    * `adj_shift_negatives_local()` same as the global method except the most
      negative value is found and adjusted on a per-channel basis.
       

## Installation

This package can be installed from CRAN

``` r
install.packages('picohdr')
```

You can install the latest development version from
[GitHub](https://github.com/coolbutuseless/picohdr) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/picohdr')
```

Pre-built source/binary versions can also be installed from
[R-universe](https://r-universe.dev)

``` r
install.packages('picohdr', repos = c('https://coolbutuseless.r-universe.dev', 'https://cloud.r-project.org'))
```

## PFM

Load and display a *Portable Float Map* (`PFM`) image.

Apply tone-mapping to the image and adjust the gamma correction prior to display.

```{r pfm1}
filename <- "man/figures/memorial.pfm.bz2"
im <- read_pfm(filename)
dim(im)

im |> 
  tm_reinhard() |>
  adj_gamma() |> 
  plot()
```



## EXR 

In this section

* Load an EXR image
* Show the names of the channels in this image
* Plot the RGB channels (apply tone-mapping and gamma adjustment)
* Plot some other channels in the image
* Display the metainformation about an EXR image

```{r exr1}
library(picohdr)

# EXR file of meta-information about the rendered scene
filename <- system.file("image/rstats.exr", package = "picohdr") 

# Load all images
images <- read_exr(filename)
dim(im)

# Channel names. EXR format wants channels arranged alphabetically
dimnames(images)[[3]]

# Extract RGB channels. Tone-map. Adjust gamma.
images[,,c('R', 'G', 'B')] |>
  tm_reinhard() |>
  adj_gamma() |>
  plot()

# Plot the albedo Green channel
plot(images[, , 'Albedo.G'])
```

### Display a non-colour channel

This file includes information about the surface derivative at each point in 
the image in the `dzdx` channel.

This value may be negative or positive, so we will map the values into the
standard range [0, 1] so it can be visualised.

```{r exr2}
# Rescale the derivative channel to the range [0,1] and display
images[,,'dzdx'] |>
  adj_rescale(0, 1) |> 
  plot()
```

### EXR metadata


```{r}
exr_info(filename)
```




```{r eval=FALSE, echo=FALSE}
im <- images[,,c('R', 'G', 'B')]
write_pfm(im, "inst/image/rstats.pfm.bz2")

im <- images[,,'G']
write_pfm(im, "inst/image/rstats-gray.pfm.bz2")
```



## EXR Implementation Notes

The code for handling EXR images is mostly written in base R with the exception 
of a core compression/decompression component which is written in C:

The ZIP compression mode applies a predictor and de-interleaves bytes. The byte 
predictor and interleaving can be in R but is 20x faster in C. 

EXR support is for a subset of images which:

* consist of a single-part scanline image.
* use the `NONE`, `ZIP` or `ZIPS` compression

This package **does not** support the following EXR features: 

* multi-part images
* deep images
* tiled images

If you would like support for these features please file an issue on GitHub.
A link to a free/openly-licensed image containing your requested features would be 
appreciated.




